Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-05-10 Manuscript/PLoSONEresponse/response.tex

The reviewer is of course correct in that real data rarely contain invariable morphological characters, because these don't contribute to phylogenetic signal. Our simulated data, on the other hand, will occasionally contain invariable characters merely because of missing data, therefore we made the decision to describe the model for morphological characters evolution as the M\textit{k} evolutionary model ($\mu$ $\geq$ 0 rather than the M\textit{kv} model where $\mu$ $>$ 0). RAxML and MrBayes did however corrected for acquisition bias as implemented in their algorithm for treating morphological data.

However, following the reviewer comment, we changed the designation of the model from M\textit{k} to M\textit{kv} to implicitly underline that we corrected for acquisition bias. We also explicitly mentioned it in the revision in the following lines:

Lines @@@112 to @@@114 in our revision:

\hfill\begin{minipage}{\dimexpr\textwidth-1cm}
We used RAxML because it automatically corrects for acquisition bias (Lewis 2001), it also is heavily used in the literature for Maximum Likelihood tree inference (e.g. [...]) and is one of the fastest methods available (Stamatakis 2008).
\end{minipage}

@@

Our simulations were designed to test the effects of missing data in as realistic a manner as possible given our constraints. Thus, we do not feel that this additional analysis would add much to the paper, especially since it is already rather long and complex.

%-------------

\textcolor{blue}{12-Filtering the data sets for strong bootstrap support under the full data analysis surely introduces some bias in favor of the no-missing data analyses.}

We implemented this step of keeping trees with a minimum total median bootstrap to make the topological comparisons more robust. If the no-missing data tree has low node support, it is likely that the changes in topology for the missing data trees are not linked to our focal question (the effect of removing characters, living taxa or fossils) but just to the low supported nodes. In fact, because we repeat the tree inference 125 times ($5^3$ parameters combinations), it is likely to have 125 different pseudo-random solution for the nodes with really low phylogentic support and that, therefore, the observed changes in topologies are just linked to the pseudo-random replication. Note however that we did not exculde this scenario from our protocol the trees selected with a median bootstrap $>$ 50 could well contain some unresolved nodes with a low bootstrap ($<$ 50). We've added the following justification in our revision:

@@

See our response to comment B above.

%-------------

\textcolor{blue}{14-line 254: RF measures the difference between the number of clades and the twice number of shared clades across two trees. The statement that it ``measures the number of shared clades across two trees" should be revised.}

We revised this statement lines @@@112 to @@@114 in our revision:

@@

\hfill\begin{minipage}{\dimexpr\textwidth-1cm}
This metric is bounded between zero, when the two trees are identical, and $2(n-2)$ (for two trees with $n$ taxa) when there is no shared clade in the two trees.
\end{minipage}

%-------------

\textcolor{blue}{16-The ``scaling" to produce the Normalized RF ``distance" is confusing. It is also inverting the sign of the distance, so it has become a measure of similarity.}

We agree that this is confusing and thank the reviewer for pointing this inconsistency out. We changed the mentions of ``Normalised Robinson-Foulds distance'' and ``Normalised Triplets distance'' to ``Normalised Robinson-Foulds metric'' and ``Normalised Triplets metric'' throughout the our revision. Additionally, we clarified the scaling method as follows (lines @@@112 to @@@114 in our revision):

@@

\hfill\begin{minipage}{\dimexpr\textwidth-1cm}
We normalised this metric following Bogdanowicz \textit{et al.} (2012)'s Normalised Tree Similarity (NTS) method. This methods scales %not sure if scale here?
any tree comparison metric using the mean distance between 1000 random trees (see Supporting Information S2: Tree Comparisons for the calculation details). This method is a generalisation of the topological accuracy method (Price \textit{et al.} 2010) allowing to compare topological differences between any tree with any tree comparison metric. In practice when the Normalised Robinson-Foulds metric between two trees is equal to one, the trees are identical; if the metric is equal to zero, the trees are no more different than expected by chance; finally if the metric is less than zero, the trees are more different than expected by chance.
\end{minipage}
% Note that NRF distance is actually a similarity ...

%-------------

\textcolor{blue}{17-line 275: why isn't the upper bound of the triples distance n choose 3 (instead of n choose 4). I think you got the formula from an unrooted quartet distance.}

We corrected this error lines @@@112 to @@@114 in our revision:

@@

\textcolor{blue}{21-line 398: ``mMthods" $->$ ``Methods"}

We thanks the reviewer for pointing out this typo and fixed it.

\textcolor{blue}{22-line 420: contra the statement here, a consensus collapsing to a polytomy will affect the RF distance (could go up or down, but it will change).}

We revised this statement at lines @@@ to @@@ in our revision

\hfill\begin{minipage}{\dimexpr\textwidth-1cm}
In this case, the Normalised Robinson-Foulds metric will decreases when the fossil are present at a lower taxonomic level (i.e. separated with more nodes from the root) but affects the clade conservation less at higher taxonomic level (i.e. separated with less nodes from the root). On the opposite though, because the fossil in a high level clade has many chances to branch on different nodes within that clade, it will be more likely to act as a wildcard taxa and decrease the Normalised Triplets metric. Therefore, the $M_{F}$ parameter is likely to affect the Normalised Robinson-Foulds metric less than the Normalised Triplets metric for the Bayesian consensus trees.
\end{minipage}

%-------------
\textcolor{blue}{Supplement 2:}

@@

\textcolor{blue}{24-The definition of the scaled RF implies that it is never positive (sinc the numerator cannot be positive and the denominator is positive).}

See our response to the comment D above.
%-------------

\textcolor{blue}{25-The decision to use Yule trees as the base line for the mean distance is not explained, and seems odd given that the true tree are not Yule trees. I'm not sure that it makes much difference, but it is odd.}

We followed the protocol in the cited reference (Bogdanowicz 2012) where the normalisation of the metrics is done on the mean value of that metric when comparing to random trees. Using random Yule trees is the best way to generate purely random trees since they are not based on any evolutionary scenario and takes only one parameter which is the number of tips (e.g. as opposed to the Birth-Death tree that assumes more realistic evolutionary scenario and takes multiple parameters).
%-------------

### 2015-05-11 Manuscript/PLoSONEresponse/TEM_PLoS_revision.tex
Minor fixed forgotten from the response

\subsection*{1. Generating the matrix}
\label{Generating_the_matrix}
First we randomly generated a ``true" tree of 50 taxa in R v. 3.0.2 \cite{R302} using the package diversitree v. 0.9-6 \cite{fitzjohndiversitree2012}. We generated the tree using a birth death process by sampling speciation ($\lambda$) and extinction ($\mu$) rates from a uniform distribution but maintaining $\lambda$ $>$ $\mu$ \cite{paradistime-dependent2011}. Empirical Total Evidence matrices vary in whether they have more fossil than living taxa or vice versa. For example, fossil taxa make up 88\% \cite{beckancient2014}, 58\% \cite{schragocombining2013}, 48\% \cite{pyrondivergence2011}, 31\% \cite{ronquista2012} and 31\% \cite{slaterphylogenetic2013} of taxa in various studies. To avoid biasing our simulations towards either living or fossil taxa and to make each simulation comparable, we implemented a rejection sampling algorithm to select only trees with 25 living and 25 fossil taxa. We then added an outgroup to the tree, using the mean branch length of the tree to separate the outgroup from the rest of the taxa, and with the branch length leading to the outgroup set as the sum of the mean branch length and the longest root-to-tip length of the tree.

Next, we generated a molecular and a morphological matrix from the ``true" tree. The molecular matrix was inferred from the ``true" tree using the R package phyclust v. 0.1-14 \cite{chen2011}. The matrix contained 1000 character sites for 51 taxa and was generated using the seqgen algorithm \cite{ranbaut1997seqgen} and using the HKY model \cite{HKY85} with random base frequencies and transition/transversion rate of two \cite{douadycomparison2003}. The substitution rates were distributed following a gamma distribution with an alpha ($\alpha$) shape of 0.5 \cite{yangamong-site1996}. We chose a low value of $\alpha$ to reduce the number of sites with high substitution rates, thus avoiding too much homoplasy and a decrease in phylogenetic signal. We selected the parameters above to generate data with no special assumption about how the characters evolved, and to reduce the computational time required if these parameters were estimated rather than defined in the tree building part of the analysis (even with the parameters defined, total computational time for the whole analysis was around 150 CPU years). All the molecular information for fossil taxa was replaced by missing data ("?").

@@

\paragraph*{Robinson-Foulds distance}
The Robinson-Foulds distance \cite{RF1981}, or ``path difference", measures the difference between the number of clades and the twice number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees \cite{RF1981} (see ~\nameref{S2_Appendix2} for calculation details). This metric is bounded between zero, when the two trees are identical, and $2(n-2)$ (for two trees with $n$ taxa) when there is no shared clade in the two trees. This metric is sensitive to minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity. We normalised this distance following Bogdanowicz \textit{et al.} \cite{Bogdanowicz2012}'s Normalised Tree Similarity (NTS) method. This methods scales %not sure if scale here?
any tree comparison metric using the mean distance between 1000 random trees (see ~\nameref{S2_Appendix2} for the calculation details). This method is a generalisation of the topological accuracy method \cite{Price2010} allowing to compare topological differences between any tree with any tree comparison metric. In practice when the Normalised Robinson-Foulds metric between two trees is equal to one, the trees are identical; if the metric is equal to zero, the trees are no more different than expected by chance; finally if the metric is less than zero, the trees are more different than expected by chance.

@@

We measured similarities among the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets metric) using the Bhattacharyya Coefficient \cite{Bhattacharyya}. The Bhattacharyya Coefficient is the probability of overlap between two distributions \cite{Bhattacharyya} (see ~\nameref{S2_Appendix2} for calculation details). Note that this is equivalent to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing distributions not means. To assess the effect of our missing data parameters, we calculated the Bhattacharyya Coefficient between the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets metric) for each pairwise combination of missing data parameters ($M_{L}$, $M_{F}$, $M_{C}$) and parameter states (0\%, 10\%, 25\%, 50\%, 75\%), i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%; $M_{L}$ = 10\%, $M_{F}$ = 0\%, $M_{C}$ = 0\% etc. (see Fig.~\ref{Fig_Bhattacharyya_Coefficients1} for more details). This resulted in 7875 pairwise comparisons (a triangular matrix with $3^5$$\times$$3^5$ cells). We performed this procedure separately for each tree inference and uncertainty method. When two combinations of missing data parameters have a similar ability to recover the ``best" tree topology the Bhattacharyya Coefficient will be close to one. Conversely, if the two combinations of missing data parameters differ, the Bhattacharyya Coefficient will be close to zero.

>>>>

\subsection*{1. Generating the matrix}
\label{Generating_the_matrix}
First we randomly generated a ``true" tree of 50 taxa in R v. 3.0.2 \cite{R302} using the package diversitree v. 0.9-6 \cite{fitzjohndiversitree2012}. We generated the tree using a birth death process by sampling speciation ($\lambda$) and extinction ($\mu$) rates from a uniform distribution but maintaining $\lambda$ $>$ $\mu$ \cite{paradistime-dependent2011}. Empirical Total Evidence matrices vary in whether they have more fossil than living taxa or vice versa. For example, fossil taxa make up 88\% \cite{beckancient2014}, 58\% \cite{schragocombining2013}, 48\% \cite{pyrondivergence2011}, 31\% \cite{ronquista2012} and 31\% \cite{slaterphylogenetic2013} of taxa in various studies. To avoid biasing our simulations towards either living or fossil taxa and to make each simulation comparable, we implemented a rejection sampling algorithm to select only trees with 25 living and 25 fossil taxa. The fossil taxa were considered as unique tips at the end of extinct lineages. We then added an outgroup to the tree, using the mean branch length of the tree to separate the outgroup from the rest of the taxa, and with the branch length leading to the outgroup set as the sum of the mean branch length and the longest root-to-tip length of the tree.

Next, we generated a molecular and a morphological matrix from the ``true" tree. The molecular matrix was inferred from the ``true" tree using the R package phyclust v. 0.1-14 \cite{chen2011}. The matrix contained 1000 character sites for 51 taxa and was generated using the seqgen algorithm \cite{ranbaut1997seqgen} and using the HKY model \cite{HKY85} with random base frequencies (sampled from a uniform probability distribution bounded between 0 and 1 with the total frequency for the four bases equal to 1) and transition/transversion rate of two \cite{douadycomparison2003}. The substitution rates were distributed following a gamma distribution with an alpha ($\alpha$) shape of 0.5 \cite{yangamong-site1996}. We chose a low value of $\alpha$ to reduce the number of sites with high substitution rates, thus avoiding too much homoplasy and a decrease in phylogenetic signal. We selected the parameters above to generate data with no special assumption about how the characters evolved, and to reduce the computational time required if these parameters were estimated rather than defined in the tree building part of the analysis (even with the parameters defined, total computational time for the whole analysis was around 150 CPU years). All the molecular information for fossil taxa was replaced by missing data ("?").

@@

\paragraph*{Robinson-Foulds distance}
The Robinson-Foulds distance \cite{RF1981}, or ``path difference", measures the difference between the number of clades and the twice number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees \cite{RF1981} (see ~\nameref{S2_Appendix2} for calculation details). This metric is bounded between zero, when the two trees are identical, and $2(n-2)$ (for two trees with $n$ taxa) when there is no shared clade in the two trees. This metric is sensitive to minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity. We normalised this metric following Bogdanowicz's Normalised Tree Similarity (NTS) method \cite{Bogdanowicz2012}. This methods scales any tree comparison metric using the mean distance between 1000 random trees (see ~\nameref{S2_Appendix2} for the calculation details). This method is a generalisation of the topological accuracy method \cite{Price2010} allowing to compare topological differences between any tree with any tree comparison metric. In practice when the Normalised Robinson-Foulds metric between two trees is equal to one, the trees are identical; if the metric is equal to zero, the trees are no more different than expected by chance; finally if the metric is less than zero, the trees are more different than expected by chance.

@@

We measured similarities among the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets metric) using the Bhattacharyya Coefficient \cite{Bhattacharyya}. The Bhattacharyya Coefficient is the probability of overlap between two distributions \cite{Bhattacharyya} (see ~\nameref{S2_Appendix2} for calculation details). Note that this is equivalent to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing whole distributions not just their means. To assess the effect of our missing data parameters, we calculated the Bhattacharyya Coefficient between the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets metric) for each pairwise combination of missing data parameters ($M_{L}$, $M_{F}$, $M_{C}$) and parameter states (0\%, 10\%, 25\%, 50\%, 75\%), i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%; $M_{L}$ = 10\%, $M_{F}$ = 0\%, $M_{C}$ = 0\% etc. (see Fig.~\ref{Fig_Bhattacharyya_Coefficients1} for more details). This resulted in 7875 pairwise comparisons (a triangular matrix with $3^5$$\times$$3^5$ cells). We performed this procedure separately for each tree inference and uncertainty method. When two combinations of missing data parameters have a similar ability to recover the ``best" tree topology the Bhattacharyya Coefficient will be close to one. Conversely, if the two combinations of missing data parameters differ, the Bhattacharyya Coefficient will be close to zero.
